Cheminformatics: The Computer Science of Chemical Discovery

### 2011-11-16 paper.tex
fixed a typo
more edits

Novel technologies in the life sciences churn out information at an
ever increasing rate -- public data stores such as the one at the
European Bioinformatics Institute (EBI) contain on the order of 10
petabytes of biological information.  While biological information has
been publicly available for decades, the same has not been true for
chemical information, until recently. It was not until 2004 that a
large public small molecule structure repository (PubChem) became
freely available. This was followed by other databases, as discussed
later in this article. While much of the foundational algorithms of
cheminformatics have been described since the 1950s, publically
available software implementing many of these algorithms have only
been accessible in the last ten to fifteen years \cite{faulon2010}.

So why should we actually care about chemical information being made
public -- and how does this relate to the field of computer science?

Making chemical information public is important because the drugs we
take are the result of intensive and costly research, which is
facilitated by the amount of information available. The more
information is shared (a process very hard to achieve in
pharmaceutical companies), the more information is available to each
researcher, facilitating the development of novel treatments. The
relevance of computer science to this research becomes clear when we
consider the quantity and types of data available -- currently a
single data base such as PubChem contains more than 34 million
chemical structure records, along with an even larger number of
\emph{annotations} (comments on the structure that are database searchable), such as synonyms (different names for the same molecule), known targets (drugs with which the molecule is known to interact), mode of action (how the molecule reacts with others),
regulatory approval status and so on.  Simply considering the synonym
list, PubChem stores 50.1 million synonym for 19.6 million compounds.

The design of data structures for this chemical information, as well
as subsequent data mining of this information, are areas where
expertise in computer science is urgently needed. (For a recent,
detailed introduction to cheminformatics, written for computer
scientists, see \cite{brown2009}).

Cheminformatics comprises different areas, which broadly speaking can
be divided into three fields: \emph{capturing data} (using lab
notebooks or potentially using formats such as the Chemical Markup
Language for publications); \emph{storing data} (designing database
schemas, devising ontologies) and \emph{mining data} (such as for
bioactivity prediction, which might involve algorithms based on
graphs).  However, the type of information being analyzed is different
from biological information: while bioinformatics often deals with
sequences, the domain of cheminformatics is chemical structures. In
the case of the former, information can very frequently be represented
as one-dimensional strings, which are relatively easy to handle
computationally. In the case of the latter however, chemical
structures may possess rings, branches as well as multiple valid
representations of the same structure (such as \emph{tautomers}, where hydrogen atoms can be positioned in different
places of the molecules), potentially giving rise to
ambiguities. Hence, chemical structures are often reported to be more
difficult to standardize.

@@

Although graphs implicitly encode the 3D structure of a molecule
(when combined with knowledge about favored bond angles and
distances), many different low energy 3D structures, or conformers,
may be consistent with the same graph structure. In addition, the 3D
structure may have a ``secondary'' geometrical arrangement of features
(such as the presence of a right-handed helix) which cannnot be
encoded in the graph. Thus we have 3D representations
(Fig.~\ref{figure:chemical-structures}C) that make the 3D arrangement
of atoms and bonds explicit. Fig.~\ref{figure:chemical-structures}D
goes further and represents a molecular surface, encoding some
property (such as lipophilicity, the ability to dissolve in fats) that varies over the surface.

@@

From the above it becomes clear that the nature of chemical problems
is often a complex mixture of combinatorial and continuous
problems. If we oversimplify the nature of chemical structures and
their properties, we can not describe the world around us in
sufficient detail. On the other hand, if we try to encode every detail
(\emph{e.g.}, all possible 3D structures and their properties, all
metabolites of a structure in humans in different tissues), the
complexity of collecting, storing, and mining these data increases
dramatically. For the field of cheminformatics to flourish, we need
closer collaboration between chemists (or, more generally, life
scientists) and computer scientists -- where the former need to be
able to pose their problems in a way relevant for practical
applications; and where the latter are able to devise ways of
capturing, storing, and analyzing chemical data which achieve optimal
balances between the different risks in generalizable encodings and
complexity (space and time). The goals of cheminformatics are to
support better chemical decision making by 1. storing and integrating
data in maintainable ways, 2.  providing enough open standards and
tools so software engineering concepts allow applications and data to
be used across heterogeneous platforms, and 3. mining the many
chemical property spaces in a time- and space-efficient way. We will
discuss all three goals in the following sections.

@@

Our motivation for this article is to highlight a variety of topics in
cheminformatics where modern CS research could contribute, and
to present the computational and algorithmic problems
that are addressed in the field.

@@

\bibliographystyle{abbrv}
\bibliography{resubmission}

>>>>

Novel technologies in the life sciences churn out information at an
ever increasing rate -- public data stores such as the one at the
European Bioinformatics Institute (EBI) contain on the order of 10
petabytes of biological information.  While biological information has
been publicly available for decades, the same has not been true for
chemical information, until recently. It was not until 2004 that a
large public small molecule structure repository (PubChem) became
freely available. This was followed by other databases. While much of
the foundational algorithms of cheminformatics have been described
since the 1950s, publicly available software implementing many of
these algorithms have only been accessible in the last ten to fifteen
years \cite{faulon2010}.

So why is chemical information important and why should we care about
its public availability -- and how does this relate to the field of
computer science?

Given that the art and science of drug discovery is based around small
molecules, techniques and algorithms to handle, analyse and interpret
chemical structure information are a core component of the drug
discovery toolbox. Public availability of chemical information and
tools to handle it are important as drug discovery research is slow,
expensive and prone to failure. The more information available to each
researcher, facilitating the development of novel treatments, the
better the chances of avoiding the many causes of attrition. The
relevance of computer science to this research becomes clear when we
consider the quantity and types of data available -- currently a
single data base such as PubChem contains more than 34 million
chemical structure records, along with an even larger number of
\emph{annotations} (comments on the structure that are database
searchable), such as synonyms (different names for the same molecule),
known targets (drugs with which the molecule is known to interact),
mode of action (how the molecule interacts with its targets),
regulatory approval status and so on.  Simply considering the synonym
list, PubChem stores 50.1 million synonym for 19.6 million compounds.

The design of data structures to handle chemical structures and
associated information, as well as subsequent data mining are areas
where expertise in computer science is urgently needed. (For a recent,
detailed introduction to cheminformatics, written for computer
scientists, see \cite{brown2009}).

Cheminformatics comprises different areas, which broadly speaking can
be divided into three fields: \emph{capturing data} (using lab
notebooks or potentially using formats such as the Chemical Markup
Language for publications); \emph{storing data} (designing database
schemas, devising ontologies) and \emph{mining data} (such as for the
prediction of biological activities of compounds).  However, the type
of information being analyzed is different from biological
information: while bioinformatics often deals with sequences, the
domain of cheminformatics is chemical structures. In the case of the
former, information can very frequently be represented as
one-dimensional strings, which are relatively easy to handle
computationally. In the case of the latter however, chemical
structures may possess rings, branches as well as multiple valid
representations of the same structure (such as \emph{tautomers}, where
hydrogen atoms can be positioned in different places of the
molecules), potentially giving rise to ambiguities. Hence, chemical
structures are often reported to be more difficult to standardize and
algorithms dealing with them must take this into account.

@@

Although graphs implicitly encode the 3D structure of a molecule (when
combined with knowledge about favored bond angles and distances), many
different low energy 3D structures, or conformers, may be consistent
with the same graph structure. In addition, the 3D structure may have
a ``secondary'' geometrical arrangement of features (such as the
presence of a right-handed helix) which cannot be encoded in the
graph. Thus we have 3D representations
(Fig.~\ref{figure:chemical-structures}C) that make the 3D arrangement
of atoms and bonds explicit. Fig.~\ref{figure:chemical-structures}D
goes further and represents a molecular surface, encoding some
property (such as lipophilicity, the ability to dissolve in non-polar
solvents) that varies over the surface.

@@

From the above it becomes clear that the nature of chemical problems
is often a complex mixture of combinatorial and continuous
problems. 
% If we oversimplify the nature of chemical structures and
% their properties, we can not describe the world around us in
% sufficient detail. On the other hand, if we try to encode every detail
% (\emph{e.g.}, all possible 3D structures and their properties, all
% metabolites of a structure in humans in different tissues), the
% complexity of collecting, storing, and mining these data increases
% dramatically. 
For the field of cheminformatics to flourish, we need
closer collaboration between chemists (or, more generally, life
scientists) and computer scientists -- where the former need to be
able to pose their problems in a way relevant for practical
applications; and where the latter are able to devise ways of
capturing, storing, and analyzing chemical data which achieve optimal
balances between the different risks in generalizable encodings and
complexity (space and time). The goals of cheminformatics are to
support better chemical decision making by 1. storing and integrating
data in maintainable ways, 2.  providing sufficient open standards and
tools allowing applications and data to be used across heterogeneous
platforms, and 3. mining the many chemical property spaces in a time-
and space-efficient way. We will discuss all three goals in the
following sections.

@@

Our motivation for this article is to highlight a variety of topics in
cheminformatics, presenting the computational and algorithmic problems
faced in the field and how modern CS research can contribute.

@@

\bibliographystyle{abbrv}
\bibliography{paper}
